---
title: ""
subtitle: ""
format: pdf 

date-format: long
execute: 
  echo: false
  warning: false
  message: false
  eval: true
---



```{r}
#| label: load-libraries
library(tidyverse)
library(ggridges)
library(viridis)
```


```{r}
salem24 <- read_csv("salem_weather_2024.csv")
```



```{r}
salem24 <- salem24 |> 
  mutate(tmax_f = (TMAX*0.18)+32,
         tmin_f = (TMIN*0.18)+32,
         wind_mph = AWND*0.2237) |>
  drop_na()
```

## Histograms

#### Plot 1: Basic Histogram (Wind Speed)



```{r}
# create a basic histogram
p1 <- ggplot(salem24, aes(x = wind_mph)) + 
  geom_histogram()
ggsave("p1.png", p1)
```

#### Plot 2: Polished Histogram (Wind Speed)


```{r}
p2 <- ggplot(salem24, aes(x = wind_mph)) + 
  geom_histogram(binwidth = 1, 
                 boundary = 0, 
                 color = "white") + 
  scale_x_continuous(breaks = seq(0,20,by=2)) + 
  labs(title = "Distribution of Average Daily Wind Speed", 
       subtitle = "Salem, Oregon 2024", 
       x = "Wind Speed (mph)", 
       y = "Count")

ggsave("p2.png", p2)
```

#### Plot 3: Histograms Faceted by Month (Wind Speed)



```{r}
salem24 <- salem24 |>
  mutate(month = month(DATE, label = TRUE, abbr = FALSE))
```


```{r}
p3 <- ggplot(salem24, aes(x = wind_mph)) + 
  geom_histogram(aes(fill = month),
                 binwidth = 1, 
                 color = "white",
                 boundary = 0) +
  scale_x_continuous(breaks = seq(0,20,by=5)) +
  guides(fill = "none") + 
  facet_wrap(vars(month)) + 
  labs(title = "Distribution of Average Daily Wind Speed", 
       subtitle = "Salem, Oregon 2024", 
       x = "Wind Speed (mph)", 
       y = "Count")

ggsave("p3.png", p3)
```


## Density Plots

#### Plot 4: Basic Density Plot (Wind Speed)



```{r}
p4 <- ggplot(salem24, aes(x = wind_mph)) + 
  geom_density(fill = "grey")

ggsave("p4.png", p4)
```

#### Plot 5: Overlaid Density Plots (Wind Speed)



```{r}
p5 <- ggplot(salem24, aes(x = wind_mph)) + 
  geom_density(aes(fill = month), 
               alpha = 0.5)

ggsave("p5.png", p5)
```

#### Plot 6: Ridge Plots (Wind Speed)



```{r}
p6 <- ggplot(salem24, aes(x = wind_mph, y = fct_rev(month))) + 
  geom_density_ridges(aes(fill = month), 
                      alpha = 0.5,
                      quantile_lines = TRUE,
                      quantiles = 2) + 
  coord_cartesian(clip = "off") + 
  scale_x_continuous(breaks = seq(0,20,by=5)) +
  guides(fill = "none") + 
  labs(title = "Distribution of Average Daily Wind Speed", 
       subtitle = "Salem, Oregon 2024", 
       x = "Wind Speed (mph)", 
       y = "") 

ggsave("p6.png", p6)
```

#### Plot 7: Ridge Plots (High Temps)

```{r}
p7 <- ggplot(salem24, aes(x = tmax_f, y = fct_rev(month))) + 
  geom_density_ridges(aes(fill = month), 
                      alpha = 0.5,
                      quantile_lines = TRUE,
                      quantiles = 2) + 
  coord_cartesian(clip = "off") +
  guides(fill = "none") + 
  labs(title = "Distribution of Daily High Temperature", 
       subtitle = "Salem, Oregon 2024", 
       x = "High Temp (F)", 
       y = "") 

ggsave("p7.png", p7)
```

#### Plot 8: Ridge Plots (High Temps with Gradient)

```{r}
p8 <- ggplot(salem24, aes(x = tmax_f, y = fct_rev(month))) + 
  geom_density_ridges_gradient(aes(fill = after_stat(x)), 
                      alpha = 0.5,
                      quantile_lines = TRUE,
                      quantiles = 2) + 
  scale_fill_viridis_c("Temp (F)",
                       option = "plasma") + 
  coord_cartesian(clip = "off") +
  labs(title = "Distribution of Daily High Temperature", 
       subtitle = "Salem, Oregon 2024", 
       x = "High Temp (F)", 
       y = "") 
p8
ggsave("p8.png", p8)
```

#### Plot 9: Ridge Plots (High and Low Temps with Gradient)

```{r}
salem24_long <- salem24 |>
  pivot_longer(cols = c(tmin_f, tmax_f),
               names_to = "temp_type",
               values_to = "temp") |>
  mutate(temp_type = recode(temp_type,
                            tmin_f = "Low",
                            tmax_f = "High")) |>
  select(DATE, temp_type, temp, month)
```

```{r}
p9 <- ggplot(salem24_long, aes(x = temp, 
                         y = fct_rev(month),
                         fill = after_stat(x),
                         linetype = temp_type)) +
  geom_density_ridges_gradient( 
                      alpha = 0.5,
                      quantile_lines = TRUE,
                      quantiles = 2) + 
  scale_fill_viridis_c("Temp (F)",
                       option = "plasma") +
  scale_linetype("") + 
  coord_cartesian(clip = "off") +
  labs(title = "Distribution of Daily High and Low Temperatures", 
       subtitle = "Salem, Oregon 2024", 
       x = "Temperature (F)", 
       y = "") 

ggsave("p9.png", p9)
```

## Rain Cloud Plots

#### Plot 10: Rain Cloud Plot (Wind Speed)

```{r}
#| fig-height: 8
#| fig-width: 8
p <- salem24 |>
  ggplot(
    aes(x = fct_rev(month), y = wind_mph)
  ) + 
  ggdist::stat_halfeye(
    aes(color = month, fill = month),
    point_color = NA, .width = 0, #adjust = -0.25,
    #width = 0.75, 
    justification = -0.2,
    alpha = 0.7
  ) + 
  geom_boxplot(
    aes(color = month), 
    outlier.shape = NA,
    width = 0.2
  ) + 
  geom_point(
    aes(color = month), 
    shape = 21,
    size = 1.5,
    position = position_jitter(seed = 1, width = 0.05)
  ) + 
  geom_point(
    aes(fill = month), 
    color = "transparent",
    shape = 21,
    size = 1.5,
    alpha = 0.3,
    position = position_jitter(seed = 1, width = 0.05)
  ) + 
  coord_flip(xlim = c(1.2, NA), clip = "off") + 
  scale_y_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 2)
  ) + 
  guides(fill = "none", color = "none") + 
  labs(x = NULL, 
       y = "Wind Speed (mph)", 
       title = "Distribution of Wind Speeds",
       subtitle = "Salem, Oregon 2024") + 
  theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(), 
    panel.grid.major.y = element_blank(),
    axis.ticks = element_blank()
  )

p
```


